Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-01-30 agujournaltemplate.tex
Post Subhi-Comments and fixing units

% MY OWN ADDED PACKAGES TO RENDER UNITS
\usepackage{gensymb}
\usepackage{booktabs}

@@

\begin{keypoints}
\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017. Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the AMO index. 
\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher in low upwelling conditions.
\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem. The next strongest predictor is the MEI v2 index (lag of 4 months), with in-situ temperature and nitrate concentration following behind. 
\end{keypoints}

@@

        %% Overarching theme: Stability of Ecosystem faunctions, Drivers of community change

        \item "The Cariaco Basin, located ...." - general hydrography/depth, etc.

@@

        % below is from 2019 REVIEW
        Hydrographic observations (Figure 2) show strong seasonality and interannual variability in the upper 400 m in the Cariaco Basin. This is controlled by seasonal and interannual changes in the wind and in the intensity of the geostrophic Caribbean Current (Muller-Karger et al. 1989). Primary upwelling was strongest between November and May. In general, during upwelling, sea surface temperatures ranged between approximately 23\degree C and 25\degree C. The secondary upwelling that occurs between June and August is shorter ($\sim$ 5 weeks) and is approximately 1.5\degree C warmer than the primary upwelling. It is driven by Ekman transport and an intensification of the curl of the wind close to the southern coast of the Caribbean (Rueda-Roa \& Muller-Karger 2013, Rueda-Roa et al. 2018). Upwelled waters move offshore in plumes that extended as far as 50 km to more than 250 km from the coast (Muller-Karger et al. 1989, 2010). From July to October, there is strong stratification in the upper 50 m of the water column after the wind weakens and the rain and rivers deliver less saline waters (Figure 2).
        Sea surface temperatures ranged from approximately 28\degree C to 30\degree C during the period of stratification (Figure 2). The mixed-layer depth varied from less than 10 m to 35 m during the 21-year time series (Figure 3). Satellite sea surface temperature and ocean color observations show that upwelling pulsates throughout the year at the scale of days to weeks, leading to marked variation in primary production (Figure 3). CARIACO rigorously characterized the annual seasonal variation in primary productivity and stratification. The length of the time series was key in capturing interannual variations that spanned decades. Satellite sea surface temperature and field data show that upwelling was stronger between 1996 and 2002. After 2002, trade wind intensity became weaker (Taylor et al. 2012).
        The field observations, particularly those in the upper 50 m, showed what appeared to be a return to stronger upwelling conditions starting in 2014 (Figure 2a).

        \item "Here's where we come in..." Talk about how most of the analysis stopped after 2012, time series died out, but there is more data there and particularly looking at community data, we can see a return.

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around 100 m. This area is relatively productive ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below 200 m depth.

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.66° W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details of the history of sampling procedures in CARIACO, as well as sampling schemes, analytical methods, and data quality control and inter-calibration procedures, have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    For the identification of phytoplankton species by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using 100 mL sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$20 μm) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level, correcting spelling variations and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}). After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024). 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps in coverage. Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis.

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (−u, easterly Trade Winds) across the southern Caribbean Sea drives near-shore upwelling of nutrient-rich waters by Ekman transport \cite{taylor_ecosystem_2012}. We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis at the grid cell closest to the Cariaco Time Series sampling coordinates (10.4°N,64.8°W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024)

@@

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of the surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (30°S-30°N and 100°E-70°W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024). In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publically available on GitHub \url{https://github.com/ben1post/cariaco-ms2}. %Note: Repo currently still set to private
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. For phytoplankton microscopy data, individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.

@@

    \subsubsection{Clustering and NMDS}
    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth. Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}. We used the Bray distance matrix in a 2-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations. The arrows indicate the gradient of a 2-dimensional linear trend surface. 
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. %Here I want to add "significant diferences calculation" using e.g. ANOVA.
    % detailed explanation of NMDS in Winder and Hunter 2008

    \subsubsection{Gradient Forest}
    To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the 21 ° C isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$). The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. To consider the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all climate variables, as well as 3 months lags for in-situ measurements in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).

\section{Draft Results}
Our study focused on phytoplankton microscopy data from the CARIACO time series, which consists of monthly samplings taken at multiple depths between November 1995 and January 2017. We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity. In total, we used a complete set of 254 monthly data points, on which all variables were available. 
% Figure 1
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll a and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. The patterns emerging from the -u 10m wind component, the temperature and the directly related 21 ° C isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the Atlantic Multidecadal Oscillation (AMO), where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

% Figure 2
To further investigate the depth distribution and dynamics of chlorophyll biomass and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as across seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters, which during some years is below the concentration between 25 and 50 meters during the years 2000, 2005, 2007, 2008 and 2013 (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. Scaling this genus diversity by the cell count abundances, i.e. the Shannon Index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall. 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

>>>>

% MY OWN ADDED PACKAGES TO RENDER UNITS
\usepackage{gensymb}
\usepackage{booktabs}
\usepackage{siunitx}

@@

\begin{keypoints}
\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017. Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index. 
\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher in low-upwelling conditions.
\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem. The next strongest predictor is the MEI v2 index (lag of 4 months), with in-situ temperature and nitrate concentration following behind. 
\end{keypoints}

@@

    \item - Phytoplankton is important

    %“Time series of biological observations provide insights into the responses of marine ecosystems to cyclic processes, transient events, and, when maintained over many decades, changes in climate (Carstensen 2014, Henson et al. 2016). Such time series are needed for assessments on the dynamic state of the ocean.” ([Muller-Karger et al., 2019, p. 2](zotero://select/library/items/7YW7MFTU)) ([pdf](zotero://open-pdf/library/items/TTZQBAXW?page=2&annotation=R4DSZPK6))

    - Phytoplankton is important
    - Why community structure is important, taxonomic diversity (aspects of diversity) - why look at taxonomic diversity?

    - some general intro to DIVERSITY.. what is it? Why important?
    %"Whereas the positive biomass response to high nutrient and light availability is well known (Tilman 1982; Leibold 1999; Siegel et al. 2013), the biodiversity response is much more complex. First, biodiversity itself is a multifaceted construct, comprising the number of different taxa (e.g. species richness) and the equality of species contribution to the assemblage as aspects of local alpha diversity (evenness). But the change in richness is only the net effect of immigration and local extinction, while the turnover in species composition with time is an additional aspect of biodiversity. Extending to larger spatial scales, heterogeneity in space (beta diversity) and composition of the regional species pool (gamma diversity) are further biodiversity dimensions. Second, for each of these facets, all metrics are effort- and scaledependent. Thus, in contrast to other aspects of ecosystem assessments, biodiversity metrics do not qualify to set absolute target values for good or bad status and effective ecosystem management. But, the relevance of the metrics arises from analysing their temporal trends over a long time period. Third, no single variable captures even the most important aspects of community composition and change. Therefore, the analysis of phytoplankton biodiversity requires a multimetric approach." from Di Cavalho et al. 2023 (with Hillebrand)

    - why are we looking at Functional Groups?
    %"The complexity of biodiversity assessment has led to the development of other approaches addressing compositional change in plankton communities, most prominently the creation of functional groups. Functional groups are meant to comprise species that show similar responses to environmental drivers and share morphological and/or physiological characteristics. Functional group indicators have been shown to be relevant for describing community structure and biodiversity and are more comparable with other studies than species-based indicators (Mouillot et al. 2006)." from Di Cavalho et al. 2023 (with Hillebrand)

    %% Overarching theme: Stability of Ecosystem faunctions, Drivers of community change
    \item "Regime shifts + Diversity" - general intro on current findings related to "climate driven ecosystem shifts"
        - %"We find that the rate of turnover in the phytoplankton community becomes faster during this century, that is, the community structure becomes increasingly unstable in response to climate change. Combined with alterations to phytoplankton diversity, our results imply a loss of ecological resilience, with likely knock-on effects on the productivity and functioning of the marine environment." \cite{henson_future_2021}
        -

    \item "The Cariaco Basin, located ...." - general hydrography/depth, etc.
        The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. 
        A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{astor_seasonal_2003}. 
        The basin is effectively separated from the Caribbean Sea by a sill with a mean depth of about 100 m. The water below the sill is poorly ventilated, which, combined with the high surface productivity, results in permanent anoxic conditions at depths below 275 m \cite{thunell_organic_2000}.

        Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!

@@

        % below is from 2019 REVIEW
        %Hydrographic observations (Figure 2) show strong seasonality and interannual variability in the upper 400 m in the Cariaco Basin. This is controlled by seasonal and interannual changes in the wind and in the intensity of the geostrophic Caribbean Current (Muller-Karger et al. 1989). Primary upwelling was strongest between November and May. In general, during upwelling, sea surface temperatures ranged between approximately 23\degree C and 25\degree C. The secondary upwelling that occurs between June and August is shorter ($\sim$ 5 weeks) and is approximately 1.5\degree C warmer than the primary upwelling. It is driven by Ekman transport and an intensification of the curl of the wind close to the southern coast of the Caribbean (Rueda-Roa \& Muller-Karger 2013, Rueda-Roa et al. 2018). Upwelled waters move offshore in plumes that extended as far as 50 km to more than 250 km from the coast (Muller-Karger et al. 1989, 2010). From July to October, there is strong stratification in the upper 50 m of the water column after the wind weakens and the rain and rivers deliver less saline waters (Figure 2).
        %Sea surface temperatures ranged from approximately 28\degree C to 30\degree C during the period of stratification (Figure 2). The mixed-layer depth varied from less than 10 m to 35 m during the 21-year time series (Figure 3). Satellite sea surface temperature and ocean color observations show that upwelling pulsates throughout the year at the scale of days to weeks, leading to marked variation in primary production (Figure 3). CARIACO rigorously characterized the annual seasonal variation in primary productivity and stratification. The length of the time series was key in capturing interannual variations that spanned decades. Satellite sea surface temperature and field data show that upwelling was stronger between 1996 and 2002. After 2002, trade wind intensity became weaker (Taylor et al. 2012).
        The field observations, particularly those in the upper 50 m, showed what appeared to be a return to stronger upwelling conditions starting in 2014 (Figure 2a).

        %“The Caribbean experiences the most marked of the El Nino–Southern Oscillation telecon- ˜ nections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Nino–Southern Oscillation (Enfield & Mayer 1997). ˜” ([Muller-Karger et al., 2019, p. 4](zotero://select/library/items/7YW7MFTU)) ([pdf](zotero://open-pdf/library/items/TTZQBAXW?page=4&annotation=D3H7YG7Y)) link to El Nino of Cariaco/Caribbean sea!

        %“When the CARIACO time series was implemented in 1995, major changes were happening in the Atlantic. We now know that at that time, the Atlantic Multidecadal Oscillation (Kerr 2000, Enfield et al. 2001) started shifting to a positive phase, which then persisted for more than 20 years and led to oceanographic anomalies in the Atlantic that were larger than any recorded over the previous 100 years (McCarthy et al. 2015). The changes had profound biological impacts throughout the Atlantic (Alexander et al. 2014, Nye et al. 2014), including in the Cariaco Basin (Taylor et al. 2012, Pinckney et al. 2015).” (Muller-Karger et al., 2019, p. 3) Interesting aspect that the AMO changed, as apparently the main climate indices to affect the Atlantic/CARIACO

    \item "Here's where we come in..." Talk about how most of the analysis stopped after 2012, time series died out, but there is more data there and particularly looking at community data, we can see a return.
    - what has been done (Phytoplankton - Statistics)?
    - where are the gaps?

    So we would hypothesize, that if wind is driving the ecosystem, then wind would be best predictor of shits.
    Or nutrients, but if it is a climate indices, then the indices capture some variablity that is not related to wind or nutrients....

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around 100 m. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below 200 m depth.

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details of the history of sampling procedures in CARIACO, as well as sampling schemes, analytical methods, and data quality control and inter-calibration procedures, have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level, correcting spelling variations and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}). After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024). 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (more than three months) in coverage. Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis. Our study instead focused on the phytoplankton microscopy data. We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity. In total, we used a complete set of 254 monthly data points, on which all variables were available.

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives near-shore upwelling of nutrient-rich waters by Ekman transport \cite{taylor_ecosystem_2012}. We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling coordinates (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024).

@@

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of the surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024). In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publicly available on GitHub \url{https://github.com/ben1post/cariaco-ms2}. %Note: Repo currently still set to private
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. For phytoplankton microscopy data, individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.

@@

    \subsubsection{Phytoplankton Community Structure Analysis}
    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth. Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}. We used the Bray distance matrix in a 2-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations. The arrows indicate the gradient of a 2-dimensional linear trend surface. 
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. %Here I want to add "significant diferences calculation" using e.g. ANOVA.
    % detailed explanation of NMDS in Winder and Hunter 2008

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the \qty{21}{\celsius} isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$). The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. To consider the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all climate variables, as well as 3 months lags for in-situ measurements in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).

\section{Draft Results}

% Figure 1
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celcius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

% Figure 2
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008 and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall. 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.
